Classless Java

### 2015-11-22 paper/sections/Overview.tex

\begin{figure}
\centering
\includegraphics[height=4cm]{PegasusDetail.pdf}
\caption{Pegasus Example.}\label{fig:pegasus}
\end{figure}

@@

Suppose that we want to model an animal system as shown in
Fig~\ref{fig:pegasus}. \texttt{Horse} and \texttt{Bird} are subtypes
of \texttt{Animal}, with methods \texttt{run()} and \texttt{fly()},
respectively. Pegasus (one of the best known creatures in Greek
mythology) can not only \emph{run} but also \emph{fly}! This is the
place where \emph{``multiple inheritance''} is needed, because
\texttt{Pegasus} needs to obtain \texttt{fly} and \texttt{run}
functionality from both \texttt{Horse} and \texttt{Bird}.
Ignoring points, for the moment, a 
first attempt to model the animal system in Java 8 is shown next:

\lstinputlisting[linerange=6-13]{../UseMixinLombok/src/pegasusJava/Main.java}% APPLY:linerange=PEGASUS_JAVA

Note that the implementations of the methods \texttt{run}
and \texttt{fly} are defined inside interfaces, using default
methods. Moreover, because interfaces support multiple interface
inheritance, the interface for \texttt{Pegasus} can inherit behaviour
from both \texttt{Horse} and \texttt{Bird}. Although Java interfaces
do not allow instance fields, no form of state is needed so far to
model the animal system.  Therefore this example provides a simple
illustratation how a basic form of multiple inheritance, similar to
what can be done with traits, can be accomplished in Java 8.

\bruno{
Note that \texttt{Bird} can fly, therefore refines the
type of \texttt{point} from \texttt{Point2D} to
\texttt{Point3D}. 
}

@@

\lstinputlisting[linerange=17-19]{../UseMixinLombok/src/pegasusJava/Main.java}% APPLY:linerange=PEGASUS_INST

@@

\noindent There are a couple of annoyances here. Firstly, the sole
purpose of the classes is to provide a way to instantiate the
objects. Although (in this case) it takes only one line code to
provide each of those classes, this code is essentially boilerplate
code, which does not add behaviour to the system. Secondly, 
the namespace gets filled with three additional types. For example, 
both \texttt{Horse} and \texttt{HorseImpl} are needed: \texttt{Horse}
is needed because it needs to be an inferface so that \texttt{Pegasus}
can use multiple inheritance; and \texttt{HorseImpl} is needed to
provide object instantiation.

Note that, for this very simple animal system, plain Java 8 anonymous
classes can be used to avoid these problems.  We could have simply
instantiated Pegasus using:

@@

However, as we shall see, once the system gets a little more
complicated, the code for instantiation quickly becomes more 
complex and verbose (even with anonymous classes).

@@

The animal system modelled so far, is a simplified version of the
system presented in Figure~\ref{}, and still does not appear to 
justify the \mixin annotation. Lets now move on to modelling 
the complete animal system. In order to do this, animals will include 
a notion of points, representing their location in space. As we shall
see, modelling stateful components using plain Java 8 quickly becomes 
cumbersome. The \mixin annotation comes to rescue here and 
allows avoiding significant amounts of boilerplate code.

\paragraph{\Q@Point2D@: simple immutable data with fields}
Two dimensional points should keep track of their coordinates. 
The usual approach to model points in Java would be to use a class 
with fields for the coordinates. However here we will illustrate how 
points are modelled with interfaces, providing a simple 
example of how to model state with interfaces. Since Java disallows fields 
inside interfaces, the state is modelled using abstract (getter) methods:

@@

However programmers are not required to use this cumbersome syntax for
every object allocation. As programmers do, for ease or reuse, the
boring long repetitive code can be encapsulate in a method. A
generalization of the \texttt{of} static factory
method is appropriate in this case:
\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int x, int y){return new Point2D(){
    public int x(){return x;} public int y(){return y;}};}
}
\end{lstlisting}

@@

\paragraph{\texttt{with} methods in Object Interfaces}
The pattern of creating a new piece of data by reusing most
information from an old piece of data is very common
when programming with immutable data-structures; it is so common that
we decided to support it in our code generation as \Q@with@ methods,
that is:
\begin{lstlisting}
@Mixin interface Point2D { 
  int x();    int y(); // getters
  Point2D withX(int val);    Point2D withY(int val); // with methods
}
\end{lstlisting}

@@

\subsection{\Q@Bird@:  field type refinement}

\subsubsection*{\Q@Point3D@ and properties updater}
\Q@Bird@ needs a 3D location, thus first of all we define \Q@Point3D@ and
as follows:
\lstinputlisting[linerange=60-64]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=POINT3D
\marco{Fix over: with(Point must be with(Point2D
and it should be after withZ, and on more lines}
Just to give you a feeling on how much boring code \mixin is generating,  we show the expanded code without then \mixin annotation\footnote{
As you can see, the generated code is very repetitive, writing such code by hand can easily induce bugs, for example a distract programmer may
swap the arguments of one of the many calls of \Q@Point3D.of@.
}.\begin{lstlisting}
interface Point3D extends Point2D{
  Point3D withX(int val);    Point3D withY(int val);    Point3D withZ(int val);
  Point3D with(Point2D val);
  public static Point3D of(int _x, int _y, int _z){
    int x=_x;    int y=_y;    int z=_z;
    return new Point3D(){
      public int x(){return x;}    public int y(){return y;}    public int z(){return z;}
      public Point3D withX(int val){return Point3D.of(val,this.y(),this.z());}
      public Point3D withY(int val){return Point3D.of(this.x(),val,this.z());}
      public Point3D withZ(int val){return Point3D.of(this.x(),this.y(),val);}
      public Point3D with(Point2D val){
        if(val instanceof Point3D){return (Point3D)val;}
        return Point3D.of(val.x(),val.y(),this.z());
        }      };    }  }
\end{lstlisting}
Note how \Q@with-@ methods are automatically refined in their return type, so that code like 
\Q@Point3D p=Point3D.of(1,2,3); p=p.withX(42);@ will be accepted and behave as expected. If the programmer wishes to suppress this behaviour and keep the signature as it was, it is sufficient to redefine the \Q@with-@ methods in the new class repeating the old signature.
Again, the philosophy is that if the programmer provides something directly, \mixin do not touch it.
\Q@with-@ methods (functionally) update a field/property at a time.
This can be inefficient, and sometime hard to maintain.
Often we want to update many fields at the same time, for example using another object as source.
Following this idea, the method \Q@with(Point2D)@ is an example of a (functional) properties updater: 
it takes a certain type and in the current object update all field in that type that match fields in the current type.
The idea is that we want as result something that is still like \Q@this@, but modified to be as much as possible similar to the parameter.
The cast in \Q@with(Point2D)@ is trivially safe since it is guarded by an \Q@instanceof@ test. The idea is that if the parameter is a subtype of the current exact type, then we can just return the parameter, as something that is just ``more'' than \Q@this@.

\subsubsection*{Unsatisfactory class based solutions to field type refinement}

\Q@Bird@s are \Q@Animal@s, 
but while \Q@Animal@s, only need 2D locations,
\Q@Bird@s need  3D locations.
In Java if we define an animal class with a field we have a set of unsatisfactory options in front of us:
\begin{itemize}
\item Define a \Q@Point3D@ field in \Q@Animal@:
this is bad since all animals would require more that is needed, and also it may requires the programmer to predict the future, or it may require to adapt the old code to accommodate for new evolutions.
\item Define a \Q@Point2D@ field in \Q@Animal@ and:
\item Define an extra \Q@int z@ field in \Q@Bird@.
This solution is very at-hoc, requires to basically duplicate the difference between \Q@Point2D@ and \Q@Point3D@ inside of \mbox{\Q@Bird@}.
Again, there are many reasons this would be bad, the most dramatic is that it would not scale to a scenario when the programmer of \Q@Bird@ and the programmer of \Q@Point3D@ are different.
\item Redefine getters and setters in \Q@Bird@, always put \Q@Point3D@ objects in the field and cast
the value out of the \Q@Point2D@ field to \Q@Point3D@ when implementing the overridden getter.
This solution scales to the multiple programmers approach, but requires ugly casts and can be implemented in a wrong way leading to bugs.
\end{itemize}

Many readers would now think that a language extension is needed.
Instead, with our language \emph{restriction} the problem can be easily challenged\footnote{Often in programming languages ``freedom is slavery''.}.
\subsubsection*{Our approach to field type refinement}

@@

In our proposed solution,
\begin{itemize}
\item by covariant method overriding we refine the field type
\item by \emph{overloading} we define a new setter with the more precise type
\item by default method we provide an implementation for the setter called with the old signature.
\end{itemize}

@@

>>>>

\begin{figure}
\begin{tabular}{l|l}
\includegraphics[height=4cm]{PegasusDetail.pdf}\hspace{20pt} &
\begin{minipage}{8cm}
\vspace{-90pt}
\lstinputlisting[linerange=6-17]{../UseMixinLombok/src/pegasusJava/Main.java}% APPLY:linerange=PEGASUS_JAVA
\end{minipage}
\end{tabular}
\caption{The animal system. Complete animal system on the left, and code for simplified animal system on the right. }\label{fig:pegasus}
\end{figure}

@@

Suppose that we want to model an animal system as shown in
the left side of Figure~\ref{fig:pegasus}. \texttt{Horse} and \texttt{Bird} are subtypes
of \texttt{Animal}, with methods \texttt{run()} and \texttt{fly()},
respectively. Pegasus (one of the best known creatures in Greek
mythology) can not only \emph{run} but also \emph{fly}! This is the
place where \emph{``multiple inheritance''} is needed, because
\texttt{Pegasus} needs to obtain \texttt{fly} and \texttt{run}
functionality from both \texttt{Horse} and \texttt{Bird}.
Ignoring points, for the moment, a 
first attempt to model the animal system in Java 8 is given on the right-side 
of Figure~\ref{fig:pegasus}.
Note that the implementations of the methods \texttt{run}
and \texttt{fly} are defined inside interfaces, using default
methods. Moreover, because interfaces support multiple interface
inheritance, the interface for \texttt{Pegasus} can inherit behaviour
from both \texttt{Horse} and \texttt{Bird}. Although Java interfaces
do not allow instance fields, no form of state is needed so far to
model the animal system.  Therefore this example provides a simple
illustratation how a basic form of multiple inheritance, similar to
what can be done with traits, can be accomplished in Java 8.

@@

\lstinputlisting[linerange=21-23]{../UseMixinLombok/src/pegasusJava/Main.java}% APPLY:linerange=PEGASUS_INST

@@

\noindent There are a couple of annoyances here. Firstly, the sole
purpose of the classes is to provide a way to instantiate the
objects. Although (in this case) it takes only one line code to
provide each of those classes, this code is essentially boilerplate
code, which does not add behaviour to the system. Secondly, 
the namespace gets filled with three additional types. For example, 
both \texttt{Horse} and \texttt{HorseImpl} are needed: \texttt{Horse}
is needed because it needs to be an inferface so that \texttt{Pegasus}
can use multiple inheritance; and \texttt{HorseImpl} is needed to
provide object instantiation. 
Note that, for this very simple animal system, plain Java 8 anonymous
classes can be used to avoid these problems.  We could have simply
instantiated Pegasus using:

@@

\noindent However, as we shall see, once the system gets a little more
complicated, the code for instantiation quickly becomes more 
complex and verbose (even with anonymous classes).

@@

The animal system modelled so far, is a simplified version of the
system presented in the left-side of Figure~\ref{fig:pegasus}, and still does not appear to 
justify the \mixin annotation. Lets now move on to modelling 
the complete animal system. In order to do this, animals will include 
a notion of points, representing their location in space. As we shall
see, modelling stateful components using plain Java 8 quickly becomes 
cumbersome. The \mixin annotation comes to rescue here and 
allows avoiding significant amounts of boilerplate code.

\paragraph{\Q@Point2D@: simple immutable data with fields}
Two dimensional points should keep track of their coordinates. 
The usual approach to model points in Java would be to use a class 
with fields for the coordinates. However here we will illustrate how 
points are modelled with interfaces, providing a simple 
example of how to model immutable state with interfaces. Since Java disallows fields 
inside interfaces, the state is modelled using abstract (getter) methods:

@@

\noindent However programmers are not required to use this cumbersome syntax for
every object allocation. As programmers do, for ease or reuse, the
boring long repetitive code can be encapsulate in a method. A
generalization of the \texttt{of} static factory
method is appropriate in this case:
\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int x, int y){return new Point2D(){
    public int x(){return x;} public int y(){return y;}};}
}
\end{lstlisting}

@@

\paragraph{\texttt{with} methods in Object Interfaces}
The pattern of creating a new piece of data by reusing most
information from an old piece of data is very common
when programming with immutable data-structures; it is so common that
it is supported in our code generation as \Q@with@ methods,
that is:
\begin{lstlisting}
@Mixin interface Point2D { 
  int x();    int y(); // getters
  Point2D withX(int val);    Point2D withY(int val); // with methods
}
\end{lstlisting}

@@

%\subsection{\Q@Bird@:  field type refinement}

\Q@Bird@s are \Q@Animal@s, but while \Q@Animal@s only need 2D
locations, \Q@Bird@s need 3D locations. Therefore when the \texttt{Bird}
interface extends the animal interface, the notion of points needs to
be refined. Such kind of refinement usually poses a challenge
with typical class-based approaches. Fortunatelly, with object interfaces,
we are able to provide a simple and effective solution to that problem.

\paragraph{Unsatisfactory class-based solutions to field type refinement}
In Java if we define an animal class with a field we have a set of unsatisfactory options in front of us:
\begin{itemize}
\item Define a \Q@Point3D@ field in \Q@Animal@:
this is bad since all animals would require more that is needed, and also it may requires the programmer to predict the future, or it may require to adapt the old code to accommodate for new evolutions.
\item Define a \Q@Point2D@ field in \Q@Animal@ and define an extra \Q@int z@ field in \Q@Bird@.
This solution is very ad-hoc, requires to basically duplicate the difference between \Q@Point2D@ and \Q@Point3D@ inside of \mbox{\Q@Bird@}.
Again, there are many reasons this would be bad, the most dramatic is that it would not scale to a scenario when the programmer of \Q@Bird@ and the programmer of \Q@Point3D@ are different.
\item Redefine getters and setters in \Q@Bird@, always put \Q@Point3D@ objects in the field and cast
the value out of the \Q@Point2D@ field to \Q@Point3D@ when implementing the overridden getter.
This solution scales to the multiple programmers approach, but requires ugly casts and can be implemented in a wrong way leading to bugs.
\end{itemize}

Many readers would now think that a language extension is needed. Instead, with object interfaces, 
another approach is possible.
%Instead, with our language \emph{restriction} the problem can be easily challenged\footnote{Often in programming languages ``freedom is slavery''.}.

\paragraph{Field type refinement with object interfaces}
In our proposed solution,
\begin{itemize}
\item by covariant method overriding we refine the field type
\item by \emph{overloading} we define a new setter with the more precise type
\item by default method we provide an implementation for the setter called with the old signature.
\end{itemize}

@@

@@

\bruno{See how to fit this in.}
\subsubsection*{\Q@Point3D@ and properties updater}
\Q@Bird@ needs a 3D location, thus first of all we define \Q@Point3D@ and
as follows:
\lstinputlisting[linerange=60-64]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=POINT3D
\marco{Fix over: with(Point must be with(Point2D
and it should be after withZ, and on more lines}
Just to give you a feeling on how much boring code \mixin is generating,  we show the expanded code without then \mixin annotation\footnote{
As you can see, the generated code is very repetitive, writing such code by hand can easily induce bugs, for example a distract programmer may
swap the arguments of one of the many calls of \Q@Point3D.of@.
}.\begin{lstlisting}
interface Point3D extends Point2D{
  Point3D withX(int val);    Point3D withY(int val);    Point3D withZ(int val);
  Point3D with(Point2D val);
  public static Point3D of(int _x, int _y, int _z){
    int x=_x;    int y=_y;    int z=_z;
    return new Point3D(){
      public int x(){return x;}    public int y(){return y;}    public int z(){return z;}
      public Point3D withX(int val){return Point3D.of(val,this.y(),this.z());}
      public Point3D withY(int val){return Point3D.of(this.x(),val,this.z());}
      public Point3D withZ(int val){return Point3D.of(this.x(),this.y(),val);}
      public Point3D with(Point2D val){
        if(val instanceof Point3D){return (Point3D)val;}
        return Point3D.of(val.x(),val.y(),this.z());
        }      };    }  }
\end{lstlisting}
Note how \Q@with-@ methods are automatically refined in their return type, so that code like 
\Q@Point3D p=Point3D.of(1,2,3); p=p.withX(42);@ will be accepted and behave as expected. If the programmer wishes to suppress this behaviour and keep the signature as it was, it is sufficient to redefine the \Q@with-@ methods in the new class repeating the old signature.
Again, the philosophy is that if the programmer provides something directly, \mixin do not touch it.
\Q@with-@ methods (functionally) update a field/property at a time.
This can be inefficient, and sometime hard to maintain.
Often we want to update many fields at the same time, for example using another object as source.
Following this idea, the method \Q@with(Point2D)@ is an example of a (functional) properties updater: 
it takes a certain type and in the current object update all field in that type that match fields in the current type.
The idea is that we want as result something that is still like \Q@this@, but modified to be as much as possible similar to the parameter.
The cast in \Q@with(Point2D)@ is trivially safe since it is guarded by an \Q@instanceof@ test. The idea is that if the parameter is a subtype of the current exact type, then we can just return the parameter, as something that is just ``more'' than \Q@this@.
